\documentclass[11pt]{article}
\usepackage[charter]{mathdesign}
\usepackage{amsmath, verbatimbox}

% Avoids spacing issues in \texttt.
\makeatletter
\let\ttfamily\relax % avoid a spurious warning
\DeclareRobustCommand\ttfamily
        {\not@math@alphabet\ttfamily\mathtt
         \fontfamily\ttdefault\selectfont\frenchspacing}
\makeatother

\title{6.945 Final Project \\ \emph{Dwimiykwim}}
\author{Ziv Scully (\texttt{ziv@mit.edu}) \\
  Daniel Shaar (\texttt{dshaar@mit.edu})}
\date{}

\begin{document}
\maketitle


\section{Argument Inference}

We have written an interpreter for a Scheme-like language
that supports automatic inference of procedure arguments.
The core of the argument inference system is in three new special forms:
\texttt{madlab}, \texttt{madblock}, and \texttt{\textasciitilde\textasciitilde}.

The \texttt{madlab} special form is a variation of \texttt{lambda}.
Specifically, the form creates a procedure
that can take its arguments in any order.
Instead of using position to match given arguments
to the variables they are bound to,
a \texttt{madlab} specifies a predicate for each input variable
that the argument bound to that variable must satisfy.
For example,
\begin{verbbox}
(define madmap
  (madlab ((xs list?) (f procedure?))
    (map f xs)))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
is a variant of \texttt{map} that can take its arguments in either order.
We call the resulting procedures ``madlab procedures'' or simply ``madlabs''.
For most purposes, madlabs are ordinary compound procedures
that happen to have unusually flexible interfaces.
For example, if we define
\begin{verbbox}
(define (curry f . args)
  (lambda more-args
    (apply f (append args more-args))))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
then both \texttt{(curry madmap exp)} and
\texttt{(curry madmap (iota 16))} work as expected,
raising $e$ to each of the elements of an input list
and mapping an input function over the integers from 0 to 15, respectively.
Note that the argument matching is not done until the madlab is applied.
To demonstrate this, consider
\begin{verbbox}
(define silly
  (madlab ((xy (member-of '(x y)))
           (yz (member-of '(y z))))
    (list xy yz)))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
and the partial application \texttt{(curry silly 'y)},
where \texttt{member-of} has the obvious namesake meaning.
When it's applied to \texttt{'x},
the \texttt{'y} is matched with \texttt{yz},
but when it's applied to \texttt{'z},
the \texttt{'y} is matched with \texttt{xy}.

The \texttt{madblock} special form is a variation of \texttt{begin}.
Just like \texttt{begin},
it groups together a sequence of expressions,
evaluates each of them in turn,
and returns the result of the last one.
The only difference is that the result of each evaluation in the sequence
is added to an \emph{inference context},
which, as we will soon explain,
is the list of values available to \texttt{??}
(which performs inference).
The inference context is dynamically bound
(the interpreter uses MIT Scheme's \texttt{fluid-let})
to the empty list at the beginning of each \texttt{madblock},
so the inference context is empty at the start of the sequence.
Our interpreter makes it easy to refer explicitly
to values of expressions earlier in the sequence
using the \texttt{define} special form
by having \texttt{define} return the result of evaluating its body
as opposed to an unspecified value or the name it was bound to.

The \texttt{??} special form is what Dwimiykwim is all about.
It takes a madlab and any number of other expressions
and applies the madlab to the given expressions
plus any additional necessary values from the inference context.
That is, it infers what arguments to pass to the given madlab.
For instance,
\begin{verbbox}
(madblock
 (curry list 'say)
 'dont-pick-me-im-a-symbol-not-a-procedure-or-list
 "a very distracting string"
 '(1 2 5 3-sir 3)
 (?? madmap))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
returns \texttt{((say 1) (say 2) (say 5) (say 3-sir) (say 3))}.
Inference only succeeds if there is an unambiguous matching
between the madlab's predicates and the union of the given expressions
and values from the inference context
with the additional constraint that every given expression is matched.
(See Section \ref{bipartite} for details.)
For example, only the first two inferences in
\begin{verbbox}
(madblock
 (curry list 'say)
 (define xs '(1 2 5 3-sir 3))
 (?? madmap)
 (?? madmap xs)
 (?? madmap))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
will succeed.
A single procedure, \texttt{(curry list 'say)},
is in the inference context the whole time.
When the first \texttt{infer} happens,
there is also only one list, so inference succeeds.
Note, however, that the resulting list is added to the inference context.
The second \texttt{??} is explicitly passed a list,
and inference succeeds thanks to this constraint.
By the time the third \texttt{??} happens
there are three lists in the inference context,
so inference fails due to ambiguity.

Just as the body of a \texttt{lambda} with multiple expressions
desugars to a single \texttt{begin} expression,
the body of a \texttt{madlab} desugars to a single \texttt{madblock}.
Additionally, each of the argument variables is added as an expression
to the beginning of the madblock.
The effect of this is that \texttt{??}
can be used freely in the body of a \texttt{madlab}
and the arguments are automatically added to the inference context.
If for some reason we need to keep the arguments out of the context,
we can write body as a \texttt{madblock} explicitly.
There is nothing to worry about with regards to nesting
because each \texttt{madblock} starts with a fresh inference context.
In fact, the intended style is for \texttt{madblock} to be rare
and mostly invoked implicitly through \texttt{madlab}.
This is because \texttt{madlab} creates a new environment,
which is a good idea given the intended synergy with \texttt{define}.


\section{Tagging}

The first piece of this project that is intended to set groundwork
for inference and matching is a lightweight system for tagging data.
The motivation for this system is that it creates a method for
describing data with useful characteristics that can be used for
performing argument inference with tag checks as predicates.
Since tags can be arbitrarily created by the user
and are not limited like type systems usually are,
the language becomes more flexible.

Tagged data consists of a value and a list of tags,
which can be used to describe the significance of the data.
We can add more tags to data as it moves through a program,
and remove the tags if we would just like the value.
However, we can always perform default operations on tagged data
without removing the tags in advanced because procedures
untag data before processing it outside the interpreter.
This is implemented by making a procedure "tag-aware".
Being "tag aware" removes tags in logical places,
so that Scheme functions and other necessary functions
can operate as intended when the tags are not important.
For example, car and cdr should be "tag aware" because
the record type for tagged data is a list,
but we only care for the actual data being tagged.

As stated earlier, since MIT Scheme does not come
with a nice type system by default,
tags allow us to work with variables
by more generally describing their characteristics.
The primary usefulness of these tags lies in argument matching.
By giving values tags, we can match tags to tag predicates
to determine if a variable is an appropriate argument to a function.
Functions can request that certain variables have certain tags,
or more generally, that these variables satisfy arbitrary predicates.
This allows us to verify that the user is performing the intended task
by forcing him/her to think in advance about what really belongs
in the function call.

A simple example to illustrate this point with out-of-order operands
is ordering tagged operands after receiving them in the wrong order.
As a quick note, the syntax for tagging data is \textasciitilde\textasciitilde.
In this example, we tag the numbers 3 and 4 with x and y respectively.
We make a function \texttt{x-then-y} that given a some x and some y,
will make a list first containing x, then y.
This utilizes the argument matching with tag checking predicates.

\begin{verbbox}
;dwimiykwim>
(define (x-then-y (x (has-tag? 'x)) (y (has-tag? 'y)))
  (list x y))

;dwimiykwim>
(x-then-y (~~ 'y 4) (~~ 'x 3))
;=> (#(<tagged> 3 (x)) #(<tagged> 4 (y)))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}

Although this tag system does not have to be used in order to perform
the core features of Dwimiykwim - argument inference,
and out of order operations - it is convenient to utilize it
in many cases where the predicate one checks
describes a characteristic of the data.


\section{Bipartite Matching}\label{bipartite}

Inferring arguments of madlabs
reduces to the following problem in graph theory.
We are given a bipartite graph with vertex partitions $A$ and $B$
satisfying $|A| \leq |B|$
along with a subset $B^* \subseteq B$ of ``required'' vertices,
and we ask whether there exists a unique matching of size $|A|$
such that every vertex of $B^*$ is matched.
$A$ is the set of predicates of the madlab's arguments,
$B^*$ is the set of values passed in explicitly,
$B$ is the union of $B^*$ and the set of values in the inference context,
and there is an edge between $a \in A$ and $b \in B$
if and only if $b$ satisfies $a$.

This problem is quite easily solved by a variation on
the traditional maximum bipartite matching algorithm,
which we review quickly here.
We refer to vertices in $A$ as ``sources'' and vertices in $B$ as ``targets''.
Let $E$ be the edge set of the graph and $M \subseteq E$ be a matching.
We think of edges in $M$ as being oriented from $B$ to $A$
and edges in $E \setminus M$ as being oriented from $A$ to $B$.
An \emph{augmenting path} of $M$ is
a path from an unmatched source to an unmatched target
that follows only edges in $E \setminus M$ from sources to targets
and only edges in $M$ from targets to sources.
Given an augmenting path of a matching,
swapping the orientations
(that is, inserting or removing from the matching as appropriate)
of all the edges in the path
yields a larger matching:
all intermediate vertices in the path remain matched,
but the previously unmatched endpoints are now matched.
This means a maximum matching has no augmenting paths.
It is well-known that the converse also holds:
if a matching has no augmenting path, it is not maximal.
Therefore, to find a maximum matching,
we repeatedly search for augmenting paths,
flipping edge orientations when we find them,
until there are no more,
at which point the edges from $B$ to $A$ are the edges of the maximum matching.

Our problem differs from traditional maximum matching in two ways.
First, we require that some targets $B^* \subseteq B$ be matched
because we must use every argument that was passed in explicitly.
To do this, we start by finding a maximum matching $M$ of $B^*$ with $A$,
reporting failure if $|M| < |B^*|$,
and then matching $A$ with $B$ using $M$ as a starting point,
guaranteeing that all of $B^*$ will remain matched.
Second, we are concerned with whether the maximum matching is unique
because there must not be multiple ways
to match all the predicates with arguments.
To do this, given a maximum matching $M$,
for each edge $e \in M$,
we attempt to find an augmenting path of the matching $M \setminus \{e\}$
in a graph with reduced edge set $E \setminus \{e\}$,
with the additional requirement that
if $e$ was incident with a vertex $b \in B^*$
then the augmenting path must finish at $b$.
That these algorithms are valid follows from the fact that,
given a non-maximum matching $M$,
there is a maximum matching containing a vertex unmatched by $M$
only if there is an augmenting path of $M$ containing that vertex,
which is a slight strengthening of the result mentioned earlier.

Once the graph is constructed, all of this can be done in quadratic time.
Our implementation is purely functional
and makes liberal use of linear-time list procedures,
so it is slower than this by approximately another quadratic factor,
but procedures generally don't take in more than, say, 8 arguments,
and $O(8^4)$ is perfectly acceptable running time for our prototype.


\section{Debugging}\label{debugging}

In addition to constructing mechanisms for tagging,
inferring arguments in function calls, and matching function arguments,
we have created a way for the user to easily correct ambiguity errors
that arise from more than one possible way to interpret an inference.
For example, consider the simple madlab below:
\begin{verbbox}
(define num-str
 (madlab ((x number?) (y string?))
         (list y x)))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
If we were to call a madblock that had two number values and strings
in its inference context,
then upon inference, we would have an ambiguity error.
To handle this case, we put the program into debug mode,
and display the context, edges, and required args to the user.
The context is indexed, so that the user does not have to spend time
typing in every expression they would like to force in the matching.
Using this information, the user is then expected to add new required args
to settle any ambiguities.
The reason we ask for required arguments is because those will be used
in the matching and will usually eliminate multiple matchings.
Once the user enters some combination of unambiguous args,
we notify the user of the successful matching and exit debug mode.
A sample workflow for \texttt{num-str} would be:
\begin{verbbox}
(madblock
 1
 "foo"
 2
 "bar"
 (?? num-str))

=== Dwimiykwim Tawimiydkwim ===

Context:
(0 "bar")
(1 2)
(2 "foo")
(3 1)

Edges:
(x 1 3)
(y 0 2)

Required:
()

New required:
(0)
Ambiguous matching!

New required:
(0 3)
Unambiguous matching! Terminate debugging mode!
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
In this example, the user was allowed to test multiple sets of required
arguments, until one achieved a good matching.
The user would then be expected to go and correct the code
by specifying those as required arguments in the inference.
The new program would now be:
\begin{verbbox}
(madblock
 1
 "foo"
 2
 "bar
 (?? num-str 1 "bar"))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
Since the user is providing so many arguments, it would not be very useful
to use inference unless they modified the code to remove some expressions.

Under the hood of the debugger, when we encounter a bad matching,
we enter debug mode, indexing each item in the context,
and print all the information we know about the matching that had an error.
We ask the user to give us a list of newly required arguments,
verifying that this is indeed a list of numbers,
and checking using the matching function if the ambiguity resolves.
In the case where the user has not specified enough required arguments,
we ask the user for a new set of required arguments,
not saving the initial choice.
This continues until either the user exits the debugger,
or a good match is found and program terminates with an error.


\section{Demo and Discussion}

As a taste of writing a real program with Dwimiykwim
and a demonstration of how much data flow can be inferred,
we devote this section to walking through the process of a sample program.
We will start by building an evaluator for arithmetic expressions
then extend it to include Scheme-style \texttt{let} bindings.
Along the way, we'll discuss some further minor features of Dwimiykwim
that would have been distracting details in the previous exposition
but are nice to have in practice.

The main task for our arithmetic evaluator is evaluating binary operations.
(We leave generalizing to $n$-ary operations as an exercise to the reader
best tried after reading this entire section.)
As a fist step, we should be able to identify
when an expression is a binary operation
and, if it is one, extract the operation name and arguments.
Just as \texttt{(define (f x1 $\dots$ xn) $\dots$)}
desugars to a \texttt{lambda} expression,
\texttt{(define (f (x1 p1?) $\dots$ (xn pn?)) $\dots$)}
desugars to a \texttt{madlab} expression
where \texttt{x1} must satisfy \texttt{p1?} and so on.
\begin{verbbox}
(define (binop? exp)
  (if (pair? exp)
      (member (car exp) '(+ - * /))
      #f))

(define (binop-op (exp binop?))
  (car exp))

(define (binop-left (exp binop?))
  (cadr exp))

(define (binop-right (exp binop?))
  (caddr exp))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
Here we use madlabs not for disambiguating arguments
but to allow inference of the single argument
when there's only a single value in the inference context
that could possibly be a binary operation expression.

Our application procedure is straightforward
other than its use of the new \texttt{has-tag?} predicate:
\texttt{((has-tag? t) x)} is true when \texttt{x} has tag \texttt{t}.
\begin{verbbox}
(define (apply-binop (op symbol?)
                     (left (has-tag? 'left))
                     (right (has-tag? 'right)))
  ((cadr (assq op
               (list (list '+ +)
                     (list '- -)
                     (list '* *)
                     (list '/ /))))
   left
   right))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
Even though \texttt{left} and \texttt{right} have tags,
we can still apply procedures from the underlying Scheme to them.
The tags are automatically stripped from the arguments of all such procedures
with a short list of exceptions including \texttt{cons} and \texttt{list}
that allow for tagged data to exist in larger data structures.

The plan for the evaluation procedure is to dispatch to two cases
depending on whether the expression is a primitive number
or a binary operation expression.
We decide between the cases using the following madlabs,
which are again madlabs for inference rather than argument ordering purposes.
\begin{verbbox}
(define (exp? x)
  (null? (tags x)))

(define (primitive? (exp exp?))
  (number? exp))

(define (operation? (exp exp?))
  (binop? exp))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
Technical note: we can't just define \texttt{binop?} as a madlab
that only accepts an expression
because during inference we need to apply it
to many values that aren't expressions,
though this could definitely be handled more intelligently
by a future version of Dwimiykwim.
At this point it is probably clear to the reader that
we have not gone to great lengths to make all of our predicates air-tight,
but our somewhat lenient definitions of \texttt{exp?} and \texttt{binop?}
suffice for this demo.

We can now finally define our main procedure.
Recall that \texttt{??} invokes argument inference,
\verb|~~| tags its second argument with its first,
and that the body of \texttt{madlab} expressions
desugar to madblocks.
This procedure introduces \texttt{madblock-inherit},
which is a variant of \texttt{madblock} that
uses the previous inference context as a starting point
instead of an empty context.
It is probably not wise to use \texttt{madblock-inherit} widely,
but here it is mostly harmless and slightly reduces clutter.
It is still somewhat safe in that values added to the inference context
don't leak into the previous inference context.
\begin{verbbox}
(define (eval (exp exp?))
  (cond
   ((?? primitive?)
    exp)
   ((?? operation?)
    (madblock-inherit
     (~~ 'left (?? eval (?? binop-left)))
     (~~ 'right (?? eval (?? binop-right)))
     (?? binop-op)
     (?? apply-binop)))))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
Notice that we almost all of the data flow is inferred:
the only explicit arguments given are for evaluating
the left and right subexpressions,
both of which are expressions, so we need to disambiguate them somehow.
The \texttt{??} isn't necessary for those \texttt{eval} applications,
but the intended style is for essentially every application
to be an inference,
which, as we'll see shortly, makes it robust to future expansion.
A quick test shows that everything works as expected.
\begin{verbbox}
;dwimiykwim>
(eval '(- (+ 3 4) 9))
;=> -2
\end{verbbox}
\begin{quote}\theverbbox\end{quote}

Adding \texttt{let} bindings to our expression language
turns out to be surprisingly little work.
In particular,
despite the fact that the evaluation procedure
will pass around a variable context,
the two cases we defined above won't change at all!
The basic operations for the \texttt{let} expressions are below.
Given what we've seen so far, they are all straightforward.
\begin{verbbox}
(define (let? exp)
  (if (pair? exp)
      (eq? (car exp) 'let)
      #f))

(define (let-vars (exp let?))
  (map car (cadr exp)))

(define (let-vals (exp let?))
  (map cadr (cadr exp)))

(define (let-body (exp let?))
  (caddr exp))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
The story is similar for context operations with one new toy:
\verb|~~:delq| removes its first argument from the tag list of its second.
\begin{verbbox}
(define ctx?
  (has-tag? 'ctx))

(define empty-ctx
  (~~ 'ctx '()))

(define (list-of p?)
  (lambda (xs)
    (if (list? xs)
        (every p? xs)
        #f)))

(define (bind (ctx ctx?)
              (vars (list-of symbol?))
              (vals (list-of number?)))
  (~~ 'ctx (append (zip list vars vals)
                   (~~:delq 'ctx ctx))))

(define (lookup (ctx ctx?)
                (var symbol?))
  (cadr (assq var ctx)))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}

We are almost ready to define evaluation,
but first we need a new inference primitive, \texttt{??:apply}.
It is actually not a primitive at all
but defined in Dwimiykwim's standard library.
\begin{verbbox}
(define (??:apply proc . args)
  (lambda more-args
    (infer proc (append args more-args))))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
In fact, not even \texttt{??} is a primitive.
\begin{verbbox}
(define (?? proc . args)
  (infer proc args))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
The real primitive is \texttt{infer},
which acts like \texttt{??}
but takes a list instead of a variable number of arguments.
This enables the definition new inference primitives like \texttt{??:apply}.
We use the word ``primitive'' not literally
but to connote that it would be bad style to define too many of these
context-dependent procedures.

With these pieces, we define predicates for the two new cases
and a new evaluation procedure.
The fact that we used \texttt{??} for the \texttt{eval} applications earlier
means we don't need to change the first two cases.
\begin{verbbox}
(define (variable? (exp exp?))
  (symbol? exp))

(define (declaration? (exp exp?))
  (let? exp))

(define (eval (ctx ctx?)
              (exp exp?))
  (cond
   ((?? primitive?)
    exp)
   ((?? operation?)
    (madblock-inherit
     (~~ 'left (?? eval (?? binop-left)))
     (~~ 'right (?? eval (?? binop-right)))
     (?? binop-op)
     (?? apply-binop)))
   ((?? variable?)
    (?? lookup))
   ((?? declaration?)
    (madblock-inherit
     (map (??:apply eval) (?? let-vals))
     (?? let-vars)
     (eval (?? bind) (?? let-body))))))
\end{verbbox}
\begin{quote}\theverbbox\end{quote}
Once again, we have to provide very few arguments explicitly,
and once again, \texttt{eval} does what it's supposed to do.
\begin{verbbox}
;dwimiykwim>
(eval
 empty-ctx
 '(let ((x (+ 2 2))
        (y (- 6 3)))
    (+ (* x x) (* y y))))
;=> 25
\end{verbbox}
\begin{quote}\theverbbox\end{quote}

What happened throughout this demo was that we shifted work from
the top-level procedure \texttt{eval} to its helper procedures.
While this first draft of Dwimiykwim is admittedly clunky,
there is a glimmer of hope in the way that the helper procedures
self-organize into the top-level procedure
with just a few essential hints as guidance.
The approach also benefits safety somewhat
by requiring the programmer to specify very specifically
how helper procedures are used.
This information typically exists anyway in documentation
(or, worse, in a single programmer's fallible memory),
and making use of it to write the program
is a potentially rich area for further research.


\end{document}
